---
title: "Project A Data"
author: "431 Staff"
date: last-modified
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

# This material is under revision. Please ignore it for now, until new material is posted.

# What Data Will I Use?

Your Project A will use the 2023 version of the analytic data from [County Health Rankings](https://www.countyhealthrankings.org/) (**CHR 2023**). Some prior CHR data has been part of our Labs, but you'll obtain, clean and manage a new data set for Project A.

The key elements we'll use are all found in the [Rankings Data & Documentation section](https://www.countyhealthrankings.org/explore-health-rankings/rankings-data-documentation), specifically the National Data & Documentation section site for the 2023 County Health Rankings. Do not use data from previous editions of the CHR, and do not use the trends data available for Project A. Specifically, you'll need to download three files in order to get and use the data:

- [the 2023 CHR CSV Analytic Data](https://www.countyhealthrankings.org/sites/default/files/media/document/analytic_data2023_0.csv) (a .csv file)
- [the 2023 CHR CSV/SAS Analytic Data Documentation](https://www.countyhealthrankings.org/sites/default/files/media/document/2023%20Analytic%20Documentation.pdf) file (PDF), and
- [the 2023 Data Dictionary](https://www.countyhealthrankings.org/sites/default/files/media/document/2023%20Data%20Dictionary%20%28PDF%29.pdf) (PDF)

In addition, when building the Codebook in Data Task 6 below, you'll use [this CHR 2023 Measures link](https://www.countyhealthrankings.org/explore-health-rankings/county-health-rankings-measures) to obtain further details on the variables you select. 

# Developing the Data: Five Data Tasks

Obtaining and cleaning your data takes a little while, but you can start at any time. First, in broad terms, here are the five data tasks you need to complete in order to build your Project A Proposal.

1. Ingest the raw data from CHR 2023 into a tibble in R called `chr_2023_raw`, then filter these raw data to only the counties actually ranked by CHR.
2. In addition to Ohio, select five more states (from the 39 others with at least 20 ranked counties) to study, so that you wind up with at least 300 and no more than 500 counties in your study. Create a new `chr_2023` tibble that contains only the counties from your chosen six states (including Ohio).
3. Reduce your data to a selection of nine variables (4 we specify below and 5 more you'll select from a list we provide), rename them in helpful ways, clean up problems, then save this smaller tibble as `chr_2023`.
4. Create a new binary factor (categorical variable) from one of the 5 variables you selected in the previous task, and add it to `chr_2023` so it now contains 10 variables.
5. Save the resulting `chr_2023` tibble as an R data set, and you will share this new `.Rds` file with us as part of your Proposal.

# Data Task 1. Ingest the raw data into `chr_2023_raw`, then filter to ranked counties

Begin an R Project just for Project A, and create a Quarto file within that project where you will do your data development work. Working from a template, or from your own best understanding of what works well for you, start by loading the packages you'll need, including the `tidyverse` and any other packages you plan to use.

You'll use `read_csv()` to ingest your raw `.csv` file into R and call the resulting tibble **chr_2023_raw**. But as you do this, you'll need to remove the top row from the .csv within your R code. You should probably look at the raw .csv in Excel or another spreadsheet system so that you know why we need to do this. 

To accomplish this, use the `skip = 1` command within your `read_csv()`. Sample code follows:

```{r}
#| eval: false

data_url <- 
  "https://www.countyhealthrankings.org/sites/default/files/media/document/analytic_data2023_0.csv"

chr_2023_raw <- read_csv(data_url, skip = 1, guess_max = 4000,
                         show_col_types = FALSE)
```

This code should create a `chr_2023_raw` tibble with 3194 rows (observations) and 720 columns (variables). 

To complete Task 1, write code to restrict the `chr_2023_raw` data to include only the **3082** rows which have "county_ranked" values of 1, since the other rows will not be used by us in this project. 

::: {.callout-note}

We strongly encourage you to use the `filter()` function from the `dplyr` package in the `tidyverse` to include only the counties ranked by CHR.

:::

At the end of Task 1, your `chr_2023_raw` tibble should have 3082 rows (counties) and 720 columns (variables).

# Data Task 2. Filter your data to the 300-500 counties in six states (including Ohio)

Now we'll filter the `chr_2023_raw` data into a new tibble called **chr_2023** which contains the 300-500 rows (counties) you will actually study in your project. 

Your selection must include six states, including: 

- the 88 counties of Ohio, and 
- all of the counties in **five** additional US states

::: {.callout-note}

1. You will write a sentence or two in your Proposal describing the reason for your selections, so you should have one.

2. Don't forget to filter in Task 1 so that only those rows with `county_ranked` values of 1 are included. Otherwise, your counts won't match those shown in the table below.

3. The table excludes Arizona, Connecticut, Delaware, District of Columbia, Hawaii, Maine, Massachusetts, Nevada, New Hampshire,  Rhode Island and Vermont because they have fewer than 20 ranked counties.
:::

## States You Can Select

The number of counties (with `county_ranked` values of 1) in each state (specified using its two-digit FIPS code and two-letter postal abbreviation) is listed below, for your convenience.

`statecode` | `state` | ranked <br /> counties | State <br /> Name | `statecode` | `state` | ranked <br /> counties | State <br /> Name
---: | ------: | -----: | ------------- | ---: | ------: | ----: | -------------
`01` | AL | 67 | Alabama | `31` | NE | 79 | Nebraska
`02` | AK | 24 | Alaska | `34` | NJ | 21 | New Jersey
`05` | AR | 75 | Arkansas | `35` | NM | 32 | New Mexico
`06` | CA | 58 | California | `36` | NY | 62 | New York
`08` | CO | 59 | Colorado | `37` | NC | 100 | North Carolina
`12` | FL | 67 | Florida | `38` | ND | 48 | North Dakota
`13` | GA | 159 | Georgia | `39` | **OH** | **88** | **Ohio**
`16` | ID | 43 | Idaho | `40` | OK | 77 | Oklahoma
`17` | IL | 102 | Illinois | `41` | OR | 35 | Oregon
`18` | IN | 92 | Indiana | `42` | PA | 67 | Pennsylvania
`19` | IA | 99 | Iowa | `45` | SC | 46 | South Carolina
`20` | KS | 104 | Kansas | `46` | SD | 61 | South Dakota
`21` | KY | 120 | Kentucky | `47` | TN | 95 | Tennessee
`22` | LA | 64 | Louisiana | `48` | TX | 244 | Texas
`24` | MD | 24 | Maryland | `49` | UT | 28 | Utah
`26` | MI | 83 | Michigan | `51` | VA | 133 | Virginia
`27` | MN | 87 | Minnesota | `53` | WA | 39 | Washington
`28` | MS | 82 | Mississippi | `54` | WV | 55 | West Virginia
`29` | MO | 115 | Missouri | `55` | WI | 72 | Wisconsin
`30` | MT | 47 | Montana | `56` | WY | 23 | Wyoming

Select six states (including `OH`) yielding a total of 300-500 counties. 

- For example, one possible combination would be `ID` (43), `OR` (35), `MT` (47), `ND` (48) and `WA` (39) with `OH` (88) yielding exactly 300 counties.
- another possibility would be `GA` (159), `CA` (58), `FL` (67), `NC` (100) and `UT` (28) with `OH` (88) yielding exactly 500 counties.

After making your selection, and filtering the data, you should have a tibble called `chr_2023` which contains all of the 300-500 counties in your six states, and 720 columns. 

::: {.callout-note}

After reducing your tibble to your selected six states, it is very important to then **convert** the `state` variable  to a `factor` in R.

:::

# Data Task 3. Add five selected measures to four required variables and clean up `chr_2023`

Next, you will identify variables from the your existing `chr_2023` tibble describing your selected sample of counties. 

In particular, you will select exactly nine variables (columns) from the 720 you should have at the start of Task 3. These nine variables must include the following four:

Variable <br /> Name | Description
-----------: | -------------------------------------------------------------
`fipscode` | the five-digit FIPS code for the county, which will be a convenient ID variable that is distinct for each row in your tibble
`county` | the name of the county, which will be useful for labeling and identifying the counties
`state` | a multi-categorical (with 6 levels) variable of two-letter postal abbreviations for your six selected states (this should be a factor variable thanks to your making it one at the end of Task 2)
`county_ranked` | tells us whether the row should be included in our data <br /> (all rows should have `county_ranked == 1`)

This set of four variables will then be followed by five variables that you will select from among the 50 listed below.

- Each of these must be of the form `vXXX_rawvalue` (note: to select the entire group of raw value variables, you might try `select(ends_with("rawvalue"))` as part of a pipe.) 
- There are many `vXXX_rawvalue` variables in the CHR data which we don't include in the list below, usually because of substantial missing data.

## Select five of these 50 variables

As part of the selection process, you should be developing appropriate **research questions** that lead to the identification of smart measures of interest (from those available) for predictors and outcomes in our two Analyses.

You must select your five variables from the 50 variables listed below. We show them here in the order in which they appear in the raw file. The listing "v001" in this table actually refers to the variable named "**v001_rawvalue**".

```
 [1] "v001" "v002" "v036" "v042" "v037" "v009" "v011" "v133" "v070" "v132"
[11] "v049" "v134" "v045" "v085" "v004" "v088" "v005" "v050" "v155" "v168"
[21] "v023" "v024" "v044" "v140" "v135" "v125" "v136" "v067" "v147" "v127"
[31] "v060" "v139" "v083" "v143" "v169" "v151" "v063" "v177" "v178" "v156" 
[41] "v153" "v154" "v166" "v053" "v054" "v056" "v126" "v059" "v057" "v058"
```

For example, `v001_rawvalue` shows the raw values for the premature death measure. If you select this variable, it is up to you to use the documentation in the two PDF files I have linked to, as well as the information on the County Health Rankings website, to get a reasonable understanding of what the variable measures, and how it was collected.

## What will these variables become?

Each of the five variables you select should be of some interest to you on its own, in terms of either providing a health outcome of interest, or potentially providing useful information about a feature of the county that might relate to that health outcome. Be sure to read this entire Data page before selecting your variables, so you are aware of some features of the data we disclose below.

- Variable 1 will be your outcome for Analysis 1. It should be a measure describing some aspect of a community's health^[Many of the variables, but probably *not* all 50 of them, meet this requirement. It's up to you to make good choices.], rather than a demographic characteristic.
- Variable 2 will be your predictor for Analysis 1, so the relationship between variables 1 and 2 should be of interest to you.
- Variable 3 will be your outcome for Analysis 2. Like variable 1, it should describe some aspect of a community's health, rather than just its demographics.
- Variable 4 (after you categorize it, later) will be your predictor for Analysis 2. Again, the relationship between variables 3 and 4 should be of interest.
- Variable 5 is an alternate choice, appropriate to serve as a replacement for any of the other four variables in your analytic work.

## Information to help select and build a codebook

The [2023 CHR Analytic Data Documentation](https://www.countyhealthrankings.org/sites/default/files/media/document/2023%20Analytic%20Documentation.pdf) file (PDF), and the [2023 Data Dictionary](https://www.countyhealthrankings.org/sites/default/files/media/document/DataDictionary_2023.pdf) file (also PDF) are crucial here, as those are the ones that explain what the available variables mean, and how they should be labeled. In building your codebook, you will also need look up each measure you select at [this CHR 2023 Measures link](https://www.countyhealthrankings.org/2023-measures).

## How To Clean Your Selected Variables

Find each of your five selected variables on one of the lists below.

- If you plan to use the variable as quantitative, do what is suggested below as part of your data development work, rename the variable appropriately, and use that version in your codebook.
- If you plan to use the variable as your categorical predictor, you should still make the appropriate change to the original quantitative version as indicated below.

The descriptions provided in these instructions associated with each variable are available as the first (deleted in R) row in the .csv file, and are also specified in the [2023 CHR Analytic Data Documentation](https://www.countyhealthrankings.org/sites/default/files/media/document/2023%20Analytic%20Documentation.pdf) PDF file.

### Ratios That Need Converting

These variables are specified as providers per population in the raw data file. You will want to take the reciprocal (1/raw value) to rescale the results in terms of population per provider, which should ease interpretation. 

- Note that after rescaling by taking the reciprocal, you may see some counties with infinite ratios, which should then be changed to **missing** values.

Code | Variable Description
---: | :---------------------------------------------
v004 | Primary care physicians
v088 | Dentists

### Variables that should be rescaled

Two of the variables you might choose should be rescaled by division so as to more closely match the scales of other variables you'll study.

Code | Variable Description | What to do?
---: | :----------- | :---------------
v001 |	Premature death	| Divide by 100 to represent losses per 1000 population
v063 |	Median household income | Divide by 1000 to represent income in thousands of dollars

### Variables that are Proportions should be converted to Percentages

Each of the variables listed below are proportions (between 0 and 1). Before you use them in any analyses, **multiply them by 100** in your data development (using the `mutate` function) to turn their values into percentages (between 0 and 100.) We're doing this to ease your interpretation of these variables.

Code | Variable Description
---: | :--------------------------------------------
v002 |	Poor or fair health
v009 |	Adult smoking
v011 |	Adult obesity
v023 |	Unemployment
v024 |	Children in poverty
v037 |	Low birthweight
v049 |	Excessive drinking
v050 |  Mammography screening
v053 |  Proportion ages 65 and older
v054 |  Proportion Non-Hispanic Black (see note below)
v056 |  Proportion Hispanic (see note below)
v057 |	Proportion Female
v058 |	Proportion Rural
v059 |  Proportion not proficient in English
v060 |	Diabetes prevalence
v067 |	Driving alone to work
v070 |	Physical inactivity
v083 |	Limited access to healthy foods
v085 |  Uninsured
v126 |  Proportion Non-Hispanic White (see note below)
v132 |	Access to exercise opportunities
v134 |  Proportion of driving deaths with alcohol involvement
v136 |	Severe housing problems
v139 |	Food insecurity
v143 |	Insufficient sleep
v153 |	Homeownership
v154 |  Severe housing cost burden
v155 |	Flu vaccinations
v166 |  Broadband access
v168 |  High school completion
v177 |  Voter turnout
v178 |  Census Participation

- A serious look at the impact of race/ethnicity is beyond the scope of Project A. If you are interested in studying race and ethnicity and their impact on a health outcome, we strongly suggest using `v126` to incorporate this dimension as a predictor (or its inverse, 1 - `v126`), rather than including the other race/ethnicity variables, specifically `v054` and `v056`. This is because there's more variation in the `v126` data across the reported counties. 

### Variables that should be OK as is

The variables listed below should be useful as they are. Most of them are ratios, although a few are averages or indexes. The main issue for these variables is correctly specifying the units of measurement (note that the indexes don't have units.)

Code | Variable Description
---: | :---------------------------------------------
v005 |	Preventable hospital stays
v036 |	Poor physical health days
v042 |	Poor mental health days
v044 |	Income inequality
v045 |	Sexually transmitted infections
v125 |	Air pollution - particulate matter
v127 |	Premature age-adjusted mortality
v133 |	Food environment index
v135 |	Injury deaths
v140 |	Social associations
v147 |	Life expectancy
v151 |  Gender pay gap
v156 |	Traffic volume
v169 |  School funding adequacy (gap measured in dollars per pupil)

## Checking Your Variables

For each of the variables you select, as part of your analytic work, you will need to both check and display that the variable has: 

- at least **80% complete** data across the counties within each of your selected states, **and**
- at least **15 distinct** values across the complete set of counties in your sample.

# Data Task 4. Create a factor for the Task 2 predictor

Create a new categorical variables (factor) based on your fourth variable, and add this new factor to your **chr_2023** tibble. You'll also retain the original (quantitative) version of this variable, so your tibble will now have 10 variables.

1. Divide the values in your variable 4 into three groups, as follows:
    - values in the lowest 40% of your sample's observations (the low group)
    - values in the middle 20% of your sample's observations (the middle group)
    - values in the highest 40% of your sample's observations (the high group)

2. Create a new factor which has two levels, low and high (although you can use other labels if you prefer), based on your three-level categorization, and treats the rest (middle group) as missing. 
    - This should yield (from your original set of 300-500 counties) between 120 and 200 defined as low, 120-200 defined as high, and 60-100 missing values in your new factor.

3. Add the resulting (binary) factor to your **chr_2023** tibble. 

# Data Task 5. Rename variables and save the final `chr_2023` tibble

Your final **chr_2023** tibble will thus contain 300-500 rows (counties) and 10 variables. You'll save this tibble as an R data set (.Rds file) and share that file with us as part of your proposal.

In your Quarto file, you will need to present all code necessary to take the original .csv data file and wind up with your `chr_2023` tibble so that we can run that code and get the same results you do.

Your main data set for analysis then should be gathered into a tibble called **chr_2023** containing the following information in this order:

- the identifying variables for each county, specifically:
    - `fipscode` = the five-digit fips code for the state and county,
    - `county` = the name of the county
    - `state` = the postal abbreviation code for the state
- the variable `county_ranked`,
- your five selected quantitative variables
- your binary categorical variable derived from your fourth selected variable

All of these variables should be **renamed** (and also have `clean_names()` from the `janitor` package applied) so as to have descriptive and maximally helpful variable names. 

- For example, if you have decided to use as a quantitative variable something like  `v009_rawvalue`, which is about adult smoking, you should rename the variable `v009_rawvalue` to **adult_smoking** in your tibble. 
- If `v009_rawvalue` (about adult smoking) is to be your categorical factor, then you should include both the original quantitative value (renamed `adult_smoking_raw`) and your categorical variable that you'll actually use in analyses, which should be named `adult_smoking_cat`.

**Save the Tibble**: You should provide code that saves your tibble as an R data set into your R Project, with the file name **chr_2023_YOURNAME.Rds**. If you like, you can store this `.Rds` file in a `data` subdirectory within your R Project.

::: {.callout-note}

1. After your cleaning is done, each row in your `chr_2023` tibble should contain all of the counties within the 6 states you are studying, and no other counties should be included in your tibble. 
2. All rows of your `chr_2023` tibble must display a  `county_ranked` value of 1.
3. If data for some counties are missing in the raw data for one or more of your variables, then these data should be indicated as missing (using NA) in the tibble.
4. There is no need to set a seed in this process, as you are not doing anything that involves selecting a random sample.

:::


